TRIK Studio: Technical Introduction

### 2017-03-10 2017-trikStudioEnglish/trikStudioEnglish.tex
remove footnotes and links from the reference list

\section{Introduction}
\label{chapter:intro}
Current state of school education in computer science turned out quite like Seymour Papert predicted it to be. In 1967 he introduced a virtual Logo turtle, that is used to teach students programming at schools even nowadays. It is less known that Papert also used a mechanical robot turtle in his experiments, that was controlled from a computer~\cite{papert1980mindstorms}, and that made educational process much more entertaining. Today Papert's ideas are widely spread, a lot of schools are using robots to teach programming (for instance, in Russia robotics is a part of compulsory education program within the Technology course~\cite{luchin2016intro}). Several robotics educational kits are used, including Lego Mindstorms NXT, Lego Mindstorms EV3\footnote{LEGO Mindstorms homepage, URL: http://www.lego.com/en-us/mindstorms (accessed: 07.02.2017)}, TRIK\footnote{TRIK robotics platform homepage, URL: http://www.trikset.com/ (accessed: 07.02.2017)} etc.

@@

Popularity of visual languages in educational robotics is proven by a number of diagram-based programming environments in this field. The most known are Robolab~\cite{erwin2000lego}, NXT-G~\cite{kelly2010lego} and EV3-G~\cite{valk2014lego}, Scratch~\cite{resnick2009scratch} and Scratch-based environments (S4A~\cite{s4a}, mBlock~\cite{mblock}, Enchanting~\cite{enchanting}, ScrathDuino~\cite{scratchduino}, Blockly~\cite{blockly} and App Inventor~\cite{wolber2011app}, 12Blocks~\cite{12blocks}, Open Roberta~\cite{jost2014graphical}, Ardublock~\cite{ardublock}), less known are Robo PRO~\cite{chang2006incorporating} for programming robots created with fischertechnik robotics kits or Program Maker for Scribbler robots and several tools for preschoolers: Lego WeDo Software~\cite{mayerova2012pilot}, Create~\cite{cross2013visual}, Wonder~\cite{wonder}.

For the last decade educational robotics has been and still is a very promising scientific field, so in 2010s almost every major university in the world invested in development of projects in robotics. For example, in 2016 Harvard University presented the Root platform~\cite{root}, Carnegie Mellon University develops and promotes its Arts\&Bots project~\cite{cross2013visual}, MIT contributed delivering Scratch environment~\cite{resnick2009scratch} and so on. Detailed overview in Russian could be found in~\cite{mordvinov2016NONPUBLISHED}, and the author concludes there that despite the variety of tools in this field there is no single one that can satisfy the needs of all educational organizations. Vast majority of such tools implement only the basic functionality (a visual editor for diagrams and an ability to run programs on robots autonomously or in a controlled manner from a computer), but they lack more advanced tools for teaching programming. Examples of such tools could be generation of readable textual code from created diagrams (that will help students to migrate from diagram-based to code-based programs), tools for debugging a program using virtual robot simulators or embedded tools for checking correctness of created programs (which could help teachers checking students' tasks). Some tools do have some of these advanced features, but often only some of them, and most of such programming environments are proprietary and pretty expensive for lots of schools.

@@

The idea of using data-flow programming languages in robotics is not a novel one, it was adopted in almost every toolset for programming industrial and laboratory automation systems, such as LabVIEW by National Instruments~\cite{kodosky1991visual}, Simulink~\cite{dabney2004mastering} and Microsoft Robotics Developer Studio~\cite{jackson2007microsoft}. They all have powerful backends (for example, Microsoft Robotics Developer Studio is being used on MySpace social network servers~\cite{scherotter2009ccr}) but are very complex and hard to learn. Thus they are used only in higher education~(\cite{stefanovic2011labview,yi2005labview}). School experiments with LabVIEW were popular in late 1990s~(\cite{cyr1997low,portsmore1999robolab}), which resulted in adaptations (like Robolab) replacing data-flow model  with control-flow model. So, there is a gap between simplified languages suitable for education and more elegant but complex industrial languages. There are attempts to adapt a data-flow model for educational robotics, for example a research group from New Zealand proposed the RuRu~\cite{diprose2011ruru} language. Nevertheless, to the best of our knowledge, there is still no production-ready data-flow-based programming environment suitable for secondary education today.

@@

Code generation mode enables users to move from visual to textual programs. Generated code is displayed in an embedded QScintilla-based\footnote{https://riverbankcomputing.com/software/qscintilla/intro (accessed: 07.02.2017)} text editor, that provides a set of common features of a code editor: syntax highlighting, code autocompletion, undo/redo, brackets highlighting etc. TRIK Studio installation contains everything needed for building and uploading programs into a robot (a number of cross-compilers, WinSCP, Putty, etc.), and it makes compilation process and communication with robots transparent to its users.

@@

In 2D simulation mode the automatic task checking feature is available (see Section~\ref{chapter:constraintsChecker}). Task checking program is written in internal event-based textual language, files containing virtual world model and task checking program can be distributed as tasks for students. Based on this feature a remote course has been launched on Stepik platform\footnote{https://stepik.org/s/7qe3xj4Z (accessed: 07.02.2017)} containing video lectures on cybernetics and robotics, numerous small tests and more than 20 tasks on educational robotics. Each task can be downloaded, solved, checked within TRIK Studio on a student's computer, and uploaded to the server where the checking system will run it on its own tests (similar to ACM ICPC coding contests).

TRIK Studio is written in C++ using cross-platform Qt\footnote{https://www.qt.io/ru/ (accessed: 14.05.2016)} framework, there are installers for Windows, Linux and Mac OS X operating systems. Because of the fact that official Lego NXT drivers are not available on Linux and Mac OS X x64, TRIK Studio contains our own implementation for them (see Section~\ref{chapter:communications} for details). The programming environment is completely free to use, open-source and is distributed under Apache License 2.0\footnote{http://www.apache.org/licenses/LICENSE-2.0 (accessed: 07.02.2017)}.

@@

Properties of each block can be set both within the diagram editor's scene and using special property editor window. All block properties (where applicable) support computable expressions written in TRIK Studio embedded textual language -- a statically typed Lua\footnote{http://www.lua.ru/ (accessed: 07.02.2017)} dialect. The parsing module for this language is written using a parser combinator library in C++11, also created within TRIK Studio project. Type inference of the resulting abstract syntax tree is performed using slightly simplified Hindley-Milner algorithm~\cite{damas1982principal}.

@@

It worth noting that in the time of writing TRIK Studio has only experimental support for data-flow language and it is not included in the officially distributed package. The source code of the editor and all appropriate tool support is freely available\footnote{https://github.com/ZiminGrigory/qreal/tree/DFVPL (accessed: 07.02.2017)} and could be compiled with the TRIK Studio's source code.

@@

USB communication is implemented using \texttt{libusb}\footnote{http://libusb.org/ (accessed: 07.02.2017)} library, Bluetooth communication is based on \texttt{QextSerialPort}\footnote{https://github.com/qextserialport/ (accessed: 07.02.2017)} library, Wi-Fi communication is implemented over TCP and UDP protocols using \texttt{QtNetwork} library, complex communication protocols (like configuration of TRIK robots before running programs) are built upon \texttt{Qt State Machine Framework}. The communication process is run in a separate thread to prevent freezing the main user interface. Particular implementations of communication mechanisms could be extended and replaced with other components. For examples, it is done for integrating TRIK Studio with the official Lego NXT driver. It worth noting that using this driver is not the only way to communicate with Lego NXT controllers over USB, TRIK Studio contains our own implementation of such a driver. Unlike the official one our driver works over libusb on all supported operating systems.

@@

\section{Textual language parser and interpreter}
\label{chapter:parser}
TRIK Studio uses visual languages for robots programming, but arithmetic expressions, intrinsic function calls and so on are better represented with textual strings (in fact, NXT-G tries to use visual blocks even for arithmetic expressions, representing them as syntax trees, it is very inconvenient). As a textual part of a language for both visual languages TRIK Studio uses a subset of Lua 5.3\footnote{https://www.lua.org/ (accessed: 07.02.2017)}, customized for our needs. The decision to implement our own parser and interpreter of a textual language was made under these considerations:

@@

The parser and the interpreter were implemented as a separate library in QReal core, so the resulting textual language is available not only in TRIK Studio, but in all other domain-specific solutions based on QReal. Parser implementation consists of two parts --- general-purpose parser combinator library and Lua parser library implemented using parser combinators. Many projects use ANTLR\footnote{http://www.antlr.org/ (accessed: 07.02.2017)}, boost.spirit\footnote{http://boost-spirit.com/home/ (accessed: 07.02.2017)}, yacc\footnote{http://dinosaur.compilertools.net/yacc/ (accessed: 07.02.2017)} as tools for parser development, but we once again created our custom solution, mainly to avoid additional external dependencies and complication of a build process --- QReal and TRIK Studio are developed by a large community, and not everyone is happy to install additional tools and learn how to use them, especially if their work has nothing to do with syntax analysis and they are students who do not took formal languages course yet.

QReal parser combinator library supports recursive-descent parsers which are able to parse a subset of LL(1) grammars. FOLLOW($\alpha$) set is not calculated, so it limits the expressiveness of resulting grammars, but we were still able to use Lua 5.3 grammar almost as it is specified, with only minor modifications related to factorization and left recursion elimination, which shall be done anyway for LL parsers. For arithmetic expressions Precedence Climbing\footnote{http://www.engr.mun.ca/\~theo/Misc/exp\_parsing.htm (accessed: 07.02.2017)} algorithm was used, and it also required some minor alterations in Lua grammar. Custom modifications were made mainly on lexer level to allow, for example, to use '!=' for inequality in addition to '\verb|~=|' used in Lua. Here is a quick example\footnote{for full specification of our parser with many irrelevant technical details see https://github.com/qreal/qreal/tree/master/qrtext/src/lua} of production written in C++ with our library (``statement is ';' or a list of expressions, optionally followed by '=' and other list of expressions''):
\begin{verbatim}
	// stat ::= ‘;’ | explist [‘=’ explist]
	stat = (-LuaTokenTypes::semicolon 
	    | (explist 
			   & ~(-LuaTokenTypes::equals & explist)))
\end{verbatim}

@@

This functionality is also used as an automatic checker of exercises for MOOC on Stepik platform\footnote{https://stepik.org/s/7qe3xj4Z (accessed: 07.02.2017)}. Without such checker it would be impossible to provide feedback to possibly large number of students, making course much less interactive. From a technical point of view the checker is a set of scripts which launch ``headless'' interpreter (i.e. interpreter without GUI) on a correct field with correct constraints. Checker is hooked up into the Stepik infrastructure and runs in a separate Docker\footnote{https://www.docker.com/ (accessed: 07.02.2017)} container when a new solution is submitted. The checker has several fields and corresponding constraint descriptions for each task (from one to five), each solution is checked against all those fields to test that it works in different situations and is not hardcoded. Fields are hidden from users, but error reporter output does gets displayed, so students can guess what went wrong (in our opinion it is better than traditional ACM ICPC system, where only a number of incorrect test and a general type of error are reported).

@@

Technical description of the TRIK Studio robots programming system was presented. TRIK Studio now has about ten thousand users across the globe, according to Google Analytics data. Now it supports English, Russian and French languages. It is open source\footnote{https://github.com/qreal/qreal (accessed: 07.02.2017)} and has large and active community developing it. TRIK Studio has about 100K lines of code (excluding QReal core, which also has size of about 120K LOC), written in C++ with Qt. Good reception of TRIK Studio in educational robotics community may be viewed as a proof that modeling environment created with domain-specific platform (QReal in this case) can actually compete with visual programming environments written specifically for robots programming.

>>>>

\section{Introduction}
\label{chapter:intro}
Current state of school education in computer science turned out quite like Seymour Papert predicted it to be. In 1967 he introduced a virtual Logo turtle, that is used to teach students programming at schools even nowadays. It is less known that Papert also used a mechanical robot turtle in his experiments, that was controlled from a computer~\cite{papert1980mindstorms}, and that made educational process much more entertaining. Today Papert's ideas are widely spread, a lot of schools are using robots to teach programming (for instance, in Russia robotics is a part of compulsory education program within the Technology course~\cite{luchin2016intro}). Several robotics educational kits are used, including Lego Mindstorms NXT, Lego Mindstorms EV3 (http://www.lego.com/en-us/mindstorms), TRIK (http://www.trikset.com/) etc.

@@

Popularity of visual languages in educational robotics is proven by a number of diagram-based programming environments in this field. The most known are Robolab~\cite{erwin2000lego}, NXT-G~\cite{kelly2010lego} and EV3-G~\cite{valk2014lego}, Scratch~\cite{resnick2009scratch} and Scratch-based environments (S4A (http://s4a.cat/), mBlock (http://www.mblock.cc/), Enchanting (http://enchanting.robotclub.ab.ca), ScrathDuino (http://www.scratchduino.ru/), Blockly (http://blockly.ru/) and App Inventor~\cite{wolber2011app}, 12Blocks (http://onerobot.org/products/12blocks/), Open Roberta~\cite{jost2014graphical}, Ardublock (http://blog.ardublock.com/), less known are Robo PRO~\cite{chang2006incorporating} for programming robots created with fischertechnik robotics kits or Program Maker for Scribbler robots and several tools for preschoolers: Lego WeDo Software~\cite{mayerova2012pilot}, Create~\cite{cross2013visual}, Wonder (https://www.makewonder.com/apps/wonder).

For the last decade educational robotics has been and still is a very promising scientific field, so in 2010s almost every major university in the world invested in development of projects in robotics. For example, in 2016 Harvard University presented the Root platform (http://http://wyss.harvard.edu/viewpage/629/), Carnegie Mellon University develops and promotes its Arts\&Bots project~\cite{cross2013visual}, MIT contributed delivering Scratch environment~\cite{resnick2009scratch} and so on. Detailed overview in Russian could be found in~\cite{mordvinov2016NONPUBLISHED}, and the author concludes there that despite the variety of tools in this field there is no single one that can satisfy the needs of all educational organizations. Vast majority of such tools implement only the basic functionality (a visual editor for diagrams and an ability to run programs on robots autonomously or in a controlled manner from a computer), but they lack more advanced tools for teaching programming. Examples of such tools could be generation of readable textual code from created diagrams (that will help students to migrate from diagram-based to code-based programs), tools for debugging a program using virtual robot simulators or embedded tools for checking correctness of created programs (which could help teachers checking students' tasks). Some tools do have some of these advanced features, but often only some of them, and most of such programming environments are proprietary and pretty expensive for lots of schools.

@@

The idea of using data-flow programming languages in robotics is not a novel one, it was adopted in almost every toolset for programming industrial and laboratory automation systems, such as LabVIEW by National Instruments~\cite{kodosky1991visual}, Simulink~\cite{dabney2004mastering} and Microsoft Robotics Developer Studio~\cite{jackson2007microsoft}. They all have powerful backends (for example, Microsoft Robotics Developer Studio is being used on MySpace social network servers, see http://channel9.msdn.com/Shows/Communicating/CCR-at-MySpace for details) but are very complex and hard to learn. Thus they are used only in higher education~(\cite{stefanovic2011labview,yi2005labview}). School experiments with LabVIEW were popular in late 1990s~(\cite{cyr1997low,portsmore1999robolab}), which resulted in adaptations (like Robolab) replacing data-flow model  with control-flow model. So, there is a gap between simplified languages suitable for education and more elegant but complex industrial languages. There are attempts to adapt a data-flow model for educational robotics, for example a research group from New Zealand proposed the RuRu~\cite{diprose2011ruru} language. Nevertheless, to the best of our knowledge, there is still no production-ready data-flow-based programming environment suitable for secondary education today.

@@

Code generation mode enables users to move from visual to textual programs. Generated code is displayed in an embedded QScintilla-based text editor, that provides a set of common features of a code editor: syntax highlighting, code autocompletion, undo/redo, brackets highlighting etc. TRIK Studio installation contains everything needed for building and uploading programs into a robot (a number of cross-compilers, WinSCP, Putty, etc.), and it makes compilation process and communication with robots transparent to its users.

@@

In 2D simulation mode the automatic task checking feature is available (see Section~\ref{chapter:constraintsChecker}). Task checking program is written in internal event-based textual language, files containing virtual world model and task checking program can be distributed as tasks for students. Based on this feature a remote course has been launched on Stepik platform containing video lectures on cybernetics and robotics, numerous small tests and more than 20 tasks on educational robotics (see https://stepik.org/s/7qe3xj4Z). Each task can be downloaded, solved, checked within TRIK Studio on a student's computer, and uploaded to the server where the checking system will run it on its own tests (similar to ACM ICPC coding contests).

TRIK Studio is written in C++ using cross-platform Qt framework (https://www.qt.io/ru/), there are installers for Windows, Linux and Mac OS X operating systems. Because of the fact that official Lego NXT drivers are not available on Linux and Mac OS X x64, TRIK Studio contains our own implementation for them (see Section~\ref{chapter:communications} for details). The programming environment is completely free to use, open-source and is distributed under Apache License 2.0 license.

@@

Properties of each block can be set both within the diagram editor's scene and using special property editor window. All block properties (where applicable) support computable expressions written in TRIK Studio embedded textual language -- a statically typed Lua dialect. The parsing module for this language is written using a parser combinator library in C++11, also created within TRIK Studio project. Type inference of the resulting abstract syntax tree is performed using slightly simplified Hindley-Milner algorithm~\cite{damas1982principal}.

@@

It worth noting that in the time of writing TRIK Studio has only experimental support for data-flow language and it is not included in the officially distributed package. The source code of the editor and all appropriate tool support is freely available and could be compiled with the TRIK Studio's source code (for the source code see https://github.com/ZiminGrigory/qreal/tree/DFVPL).

@@

USB communication is implemented using \texttt{libusb} library (http://libusb.org/), Bluetooth communication is based on \texttt{QextSerialPort} library (https://github.com/qextserialport/), Wi-Fi communication is implemented over TCP and UDP protocols using \texttt{QtNetwork} library, complex communication protocols (like configuration of TRIK robots before running programs) are built upon \texttt{Qt State Machine Framework}. The communication process is run in a separate thread to prevent freezing the main user interface. Particular implementations of communication mechanisms could be extended and replaced with other components. For examples, it is done for integrating TRIK Studio with the official Lego NXT driver. It worth noting that using this driver is not the only way to communicate with Lego NXT controllers over USB, TRIK Studio contains our own implementation of such a driver. Unlike the official one our driver works over libusb on all supported operating systems.

@@

\section{Textual language parser and interpreter}
\label{chapter:parser}
TRIK Studio uses visual languages for robots programming, but arithmetic expressions, intrinsic function calls and so on are better represented with textual strings (in fact, NXT-G tries to use visual blocks even for arithmetic expressions, representing them as syntax trees, it is very inconvenient). As a textual part of a language for both visual languages TRIK Studio uses a subset of Lua 5.3 (https://www.lua.org/), customized for our needs. The decision to implement our own parser and interpreter of a textual language was made under these considerations:

@@

The parser and the interpreter were implemented as a separate library in QReal core, so the resulting textual language is available not only in TRIK Studio, but in all other domain-specific solutions based on QReal. Parser implementation consists of two parts --- general-purpose parser combinator library and Lua parser library implemented using parser combinators. Many projects use ANTLR (http://www.antlr.org/), boost.spirit (http://boost-spirit.com/home/), yacc (http://dinosaur.compilertools.net/yacc/) as tools for parser development, but we once again created our custom solution, mainly to avoid additional external dependencies and complication of a build process --- QReal and TRIK Studio are developed by a large community, and not everyone is happy to install additional tools and learn how to use them, especially if their work has nothing to do with syntax analysis and they are students who do not took formal languages course yet.

QReal parser combinator library supports recursive-descent parsers which are able to parse a subset of LL(1) grammars. FOLLOW($\alpha$) set is not calculated, so it limits the expressiveness of resulting grammars, but we were still able to use Lua 5.3 grammar almost as it is specified, with only minor modifications related to factorization and left recursion elimination, which shall be done anyway for LL parsers. For arithmetic expressions Precedence Climbing algorithm (see http://www.engr.mun.ca/\~theo/Misc/exp\_parsing.htm for details) was used, and it also required some minor alterations in Lua grammar. Custom modifications were made mainly on lexer level to allow, for example, to use '!=' for inequality in addition to '\verb|~=|' used in Lua. Here is a quick example of production written in C++ with our library (``statement is ';' or a list of expressions, optionally followed by '=' and other list of expressions''):
\begin{verbatim}
	// stat ::= ‘;’ | explist [‘=’ explist]
	stat = (-LuaTokenTypes::semicolon 
	    | (explist 
			   & ~(-LuaTokenTypes::equals & explist)))
\end{verbatim}

For full specification of our parser with many irrelevant technical details see https://github.com/qreal/qreal/tree/master/qrtext/src/lua.

@@

This functionality is also used as an automatic checker of exercises for MOOC on Stepik platform. Without such checker it would be impossible to provide feedback to possibly large number of students, making course much less interactive. From a technical point of view the checker is a set of scripts which launch ``headless'' interpreter (i.e. interpreter without GUI) on a correct field with correct constraints. Checker is hooked up into the Stepik infrastructure and runs in a separate Docker container when a new solution is submitted. The checker has several fields and corresponding constraint descriptions for each task (from one to five), each solution is checked against all those fields to test that it works in different situations and is not hardcoded. Fields are hidden from users, but error reporter output does gets displayed, so students can guess what went wrong (in our opinion it is better than traditional ACM ICPC system, where only a number of incorrect test and a general type of error are reported).

@@

Technical description of the TRIK Studio robots programming system was presented. TRIK Studio now has about ten thousand users across the globe, according to Google Analytics data. Now it supports English, Russian and French languages. It is open source (https://github.com/qreal/qreal) and has large and active community developing it. TRIK Studio has about 100K lines of code (excluding QReal core, which also has size of about 120K LOC), written in C++ with Qt. Good reception of TRIK Studio in educational robotics community may be viewed as a proof that modeling environment created with domain-specific platform (QReal in this case) can actually compete with visual programming environments written specifically for robots programming.
